D-Hammer: Efficient Equational Reasoning for Labelled Dirac Notation

### 2025-01-31 02language_design.tex
sound and complete for normalization

\begin{lemma}[Soundness of type system]
  If \( \Gamma \vdash t : T \), then for all valuations \( v \) valid w.r.t.\,
  $\Gamma$, we have \( \sem{t}_v \in \sem{T}_v \).
\end{lemma}
% \begin{proof}
% By induction on the derivation.
% \end{proof}
This interpretation formalizes the standard understanding of Dirac
notation and provides the foundation for the algorithm. However,
computers cannot directly reason about equivalence through
mathematical interpretations. We proceed by defining a proof system
that abstracts these concepts.

>>>>

\begin{lemma}[Soundness of type system]
  \label{lemma:sound type system}
  If \( \Gamma \vdash t : T \), then for all valuations \( v \) valid w.r.t.\,
  $\Gamma$, we have \( \sem{t}_v \in \sem{T}_v \).
\end{lemma}
% \begin{proof}
% By induction on the derivation.
% \end{proof}
This interpretation formalizes the standard understanding of Dirac
notation and provides the foundation for the algorithm. However,
computers cannot directly reason about equivalence through
mathematical interpretations. We proceed by defining a proof system
that abstracts these concepts.

### 2025-01-31 03labelled.tex
sound and complete for normalization

\paragraph*{Typing rules.}
There are various rules for computing types and checking vadility of registers and labelled terms. Here we display some of the rules and see Appendix for all the rest.
\begin{gather*}
  \frac{
      \Gamma \vdash R : \reg(\sigma) \quad
      \Gamma \vdash Q : \reg(\tau)
      \quad \var(R) \cap \var(Q) = \emptyset
  }{\Gamma \vdash (R,Q) : \reg(\sigma \times \tau)} \\[0.2cm]
  \frac{
          \Gamma \vdash r : \reg(\sigma) \quad
          \Gamma \vdash i : \Basis(\sigma)}
  {\Gamma \vdash |i\>_r : \DType(\{r\}, \emptyset)}
  \qquad
  \frac{\Gamma \vdash R : \reg(\sigma)\qquad \Gamma \vdash K : \KType(\sigma)}{\Gamma \vdash K_R : \DType(\var(R), \emptyset)} \\[0.2cm]
    \frac{
        \Gamma \vdash D_i : \DType(s_i,s_i') \qquad
        \forall\,i\neq j.\ s_i\cap s_j = \emptyset \qquad
        \forall\,i\neq j.\ s'_i\cap s'_j = \emptyset
    }
    {\Gamma \vdash D_1 \otimes \cdots \otimes D_i : \DType(\bigcup_i s_i, \bigcup_i s_i')}.
\end{gather*}
The first rule says that a paired register is of the product type and its components should be disjoint.
To lift a plain Dirac notation into the labelled one (line 2), we enforce that the term and register are of the same indices, reflecting the fact the state should be consistant to the corresponding subsystems.
The third line gives the typing of labelled tensoring with a check whether the component subsystems are disjoint with each other.

@@

\begin{definition}[Cylindrical Extension]
  For any $D : \cD(s_1,s_2)$ and $s$ that disjoint with both $s_1$ and $s_2$, we define $\cl(D,s) \triangleq D\otimes \mathbf{1}_{s}$ of type $\cD(s_1\cup s, s_2\cup s)$.
\end{definition}
Formally, we equip $\cR$ with a default order, say for example the alphabetical order of names. For any valid register $R$, there exists the operator $\Swap_R$ that sorts $R$; for example, if $\Gamma\vdash p : \reg(\sigma_p)$ and $\Gamma\vdash q : \reg(\sigma_q)$, then $\Swap_{(q,p)} \triangleq \sum_{i\in \bU(\sigma_p)}\sum_{j\in \bU(\sigma_q)}|i\>\<j|\otimes |j\>\<i|$. We can further define $\Swap_{s_1,s_2,\cdots}$ for merging disjoint sets orderly.

For given context $\Gamma$ and set $s = \{r_1,\cdots,r_n\}\subseteq\cR$ ($r_i$ is sorted) with $\Gamma\vdash r_i : \sigma_i$, we define $\sigma_s\triangleq \sigma_1\times\cdots\times\sigma_n$ if $s$ is non-empty. \label{context: sigma s}
For any valuation $v$, we further write $\cH_{\sem{s}_v} \triangleq \bigotimes_{i}\cH_{\sem{\sigma_i}_v}$ which equals to $\cH_{\sem{\sigma_s}_v}$ for non-empty $s$.
% for any $s\subseteq \cR$, where $\Gamma\vdash r : \reg(\sigma_r)$ for all $r\in\cR$. 
We interpret labelled Dirac notations inductively as (assume $\Gamma\vdash D_i : \cD(s_i,s'_i)$) :
\begin{itemize}
  \item $\sem{|i\>_r}_v = |\sem{i}_v\>$;\quad 
        $\sem{\<i|_r}_v = \<\sem{i}_v|$;\quad  
        $\sem{K_R}_v = \sem{\Swap_R}_v \cdot \sem{K}_v$; \\
        $\sem{B_R}_v = \sem{B}_v\cdot \sem{\Swap_R}_v$;\quad
        $\sem{O_{R_1,R_2}} = \sem{\Swap_{R_1}}_v\cdot \sem{O}_v\cdot \sem{\Swap_{R_2}}_v$;
  \item $\sem{D_1\otimes \cdots \otimes D_n}_v = \sem{\Swap_{s_1,\cdot,s_n}}_v\cdot (\sem{D_1}_v\otimes \cdots \otimes \sem{D_1}_v)\cdot \sem{\Swap_{s'_1,\cdot,s'_n}}_v$;
  \item $\sem{D_1\cdot D_2}_v = \sem{\cl(D_1, s_2\backslash s'_1)}_v\cdot \sem{\cl(D_2, s'_1\backslash s_2)}_v$. Note that $s_2\backslash s'_1$ and $s'_1\backslash s_2$ are the minimal extension that make it interpretable. E.g., to interpret ${}_p\<i|\cdot |j\>_{p,q}$, we at least need to extend ${}_p\<i|$ to ${}_p\<i|\otimes I_{q}$.
\end{itemize}
Thus, $\sem{D}_v$ is a linear map from $\cH_{\sem{s'}_v}$ to $\cH_{\sem{s}_v}$ given $\Gamma\vdash D : \cD(s,s')$. Following the semantics, labelled tensor is independent of its order, i.e., $\sem{D_1\otimes D_2}_v = \sem{D_2\otimes D_1}_v$ and $\sem{D_1\otimes (D_2\otimes D_3)}_v = \sem{(D_1\otimes D_2)\otimes D_3}_v$, which ensures the soundness of treating labelled tensor as an AC symbol.

@@

Define $T_{s,s'}$ as $\SType$ if $s=s'=\emptyset$, $\KType(\sigma_s)$ if $s'=\emptyset$, $\BType(\sigma_{s'})$ if $s=\emptyset$ and $\OType(\sigma_s,\sigma_{s'})$ otherwise (see Sec. \ref{context: sigma s}). We claim the soundness of our normalization as:
% Then if the original equality is $\Gamma\vdash D_1 = D_2 : \cD(s,s')$, what remains to be check $\Gamma\vdash e_1 = e_2 : T_{s,s'}$.

\begin{theorem}[Soundness and Completeness of Normalization]
  Assuming $\Gamma\vdash D_1 : \cD(s,s')$, $\Gamma\vdash D_2 : \cD(s,s')$, 
  and no variables of $\DType(\cdot,\cdot)$ in $D_1,D_2$.
  Let $e_1 = e_2$ be obtained by above normalization procedure on $D_1 = D_2$. Then:
  \begin{itemize}
    \item (Soundness) $\Gamma\vdash e_1 = e_2 : T_{s,s'}$ implies $\Gamma\vDash D_1 = D_2$;
    \item (Completeness) $\Gamma\vDash D_1 = D_2$ implies $\Gamma\vDash e_1 = e_2$
  \end{itemize}
  %All rewriting rules for labelled Dirac notations listed here and in Appendix are sound, i.e., for any interpretation $\sem{\cdot}$, $D\reduce D'$ implies $\sem{D} = \sem{D'}$.
\end{theorem}
% \begin{proof}
%   Notice that all rewriting rules and ordering perserve the types, so the final step eliminating labels is both sound and complete. 
%   It is routinely to check that each rewriting rules and ordering step are sound based on the semantics. This directly leads to the soundness. For completeness, we need further check if the normalization procedure always provide $e_1 = e_2$, i.e., it never gets stuck and terminates. Step 1 is trivial. For step 2, it can be further checked that 1. there is no circular rules and thus terminates, and 2. if it is not in form (\ref{eqn:Labelled normal}), there must exists one of the rules can be further applied. For step 3, ordering is always applicable. This completes the proof.
% \end{proof}

>>>>

\paragraph*{Typing rules.}
There are various rules for computing types and checking vadility of registers and labelled terms. Here we display some of the rules and see Appendix for all the rest.
\begin{gather*}
  \frac{
      \Gamma \vdash R : \reg(\sigma) \quad
      \Gamma \vdash Q : \reg(\tau)
      \quad \var(R) \cap \var(Q) = \emptyset
  }{\Gamma \vdash (R,Q) : \reg(\sigma \times \tau)} \\[0.2cm]
  \frac{
          \Gamma \vdash r : \reg(\sigma) \quad
          \Gamma \vdash i : \Basis(\sigma)}
  {\Gamma \vdash |i\>_r : \DType(\{r\}, \emptyset)}
  \qquad
  \frac{\Gamma \vdash R : \reg(\sigma)\qquad \Gamma \vdash K : \KType(\sigma)}{\Gamma \vdash K_R : \DType(\var(R), \emptyset)} \\[0.2cm]
    \frac{
        \Gamma \vdash D_i : \DType(s_i,s_i') \qquad
        \forall\,i\neq j.\ s_i\cap s_j = \emptyset \qquad
        \forall\,i\neq j.\ s'_i\cap s'_j = \emptyset
    }
    {\Gamma \vdash D_1 \otimes \cdots \otimes D_i : \DType(\bigcup_i s_i, \bigcup_i s_i')}.
\end{gather*}
The first rule says that a paired register is of the product type and its components should be disjoint.
To lift a plain Dirac notation into the labelled one (line 2), we enforce that the term and register are of the same indices, reflecting the fact the state should be consistant to the corresponding subsystems.
The third line gives the typing of labelled tensoring with a check whether the component subsystems are disjoint with each other.
Since $\cR$ is given, let $\sigma_r : \Index$ denote type of $r$, i.e.,  $\Gamma\vdash r : \reg(\sigma_r)$, for simplicity.

@@

\begin{definition}[Cylindrical Extension]
  For any $D : \cD(s_1,s_2)$ and $s$ that disjoint with both $s_1$ and $s_2$, we define $\cl(D,s) \triangleq D\otimes \mathbf{1}_{s}$ of type $\cD(s_1\cup s, s_2\cup s)$.
\end{definition}
Formally, we equip $\cR$ with a default order, say for example the alphabetical order of names. For any valid register $R$, there exists the operator $\Swap_R$ that sorts $R$; for example, $\Swap_{(q,p)} \triangleq \sum_{i\in \bU(\sigma_p)}\sum_{j\in \bU(\sigma_q)}|i\>\<j|\otimes |j\>\<i|$. We can further define $\Swap_{s_1,s_2,\cdots}$ for merging disjoint sets orderly.

For given context $\Gamma$ and any valuation $v$,
we interpret 
$$\mbox{$\sem{\cD(s,s')} \triangleq \mathcal{L}(\bigotimes_{j}\cH_{\sem{\sigma_{r'_j}}_v}, \bigotimes_{i}\cH_{\sem{\sigma_{r_i}}_v})$}$$ where $\mathcal{L}$ denotes the set of linear maps, 
the sets $s = \{r_1,\cdots,r_n\}$ and $s' = \{r'_1,\cdots,r'_m\}$ ($r_i$ and $r'_j$ are sorted).
% we define $\sigma_s\triangleq \sigma_1\times\cdots\times\sigma_n$ if $s$ is non-empty. \label{context: sigma s}
% For any valuation $v$, we further write $\cH_{\sem{s}_v} \triangleq \bigotimes_{i}\cH_{\sem{\sigma_i}_v}$ which equals to $\cH_{\sem{\sigma_s}_v}$ for non-empty $s$.
% We interpret $\sem{\cD(s,s')} \triangleq \mathcal{L}(\cH_{\sem{s'}_v}, \cH_{\sem{s}_v})$ where $\mathcal{L}$ denotes the set of linear maps.
% for any $s\subseteq \cR$, where $\Gamma\vdash r : \reg(\sigma_r)$ for all $r\in\cR$. 
We interpret labelled Dirac notations inductively as (assume $\Gamma\vdash D_i : \cD(s_i,s'_i)$) :
\begin{itemize}
  \item $\sem{|i\>_r}_v = |\sem{i}_v\>$;\quad 
        $\sem{\<i|_r}_v = \<\sem{i}_v|$;\quad  
        $\sem{K_R}_v = \sem{\Swap_R}_v \cdot \sem{K}_v$; \\
        $\sem{B_R}_v = \sem{B}_v\cdot \sem{\Swap_R}_v$;\quad
        $\sem{O_{R_1,R_2}} = \sem{\Swap_{R_1}}_v\cdot \sem{O}_v\cdot \sem{\Swap_{R_2}}_v$;
  \item $\sem{D_1\otimes \cdots \otimes D_n}_v = \sem{\Swap_{s_1,\cdot,s_n}}_v\cdot (\sem{D_1}_v\otimes \cdots \otimes \sem{D_1}_v)\cdot \sem{\Swap_{s'_1,\cdot,s'_n}}_v$;
  \item $\sem{D_1\cdot D_2}_v = \sem{\cl(D_1, s_2\backslash s'_1)}_v\cdot \sem{\cl(D_2, s'_1\backslash s_2)}_v$. Note that $s_2\backslash s'_1$ and $s'_1\backslash s_2$ are the minimal extension that make it interpretable. E.g., to interpret ${}_p\<i|\cdot |j\>_{p,q}$, we at least need to extend ${}_p\<i|$ to ${}_p\<i|\otimes I_{q}$.
\end{itemize}
It can be shown that Lemma \ref{lemma:sound type system} also holds for labelled terms, i.e., $\sem{D}_v \in \sem{\cD(s,s')}_v$ given $\Gamma\vdash D : \cD(s,s')$. Following the semantics, labelled tensor is independent of its order, i.e., $\sem{D_1\otimes D_2}_v = \sem{D_2\otimes D_1}_v$ and $\sem{D_1\otimes (D_2\otimes D_3)}_v = \sem{(D_1\otimes D_2)\otimes D_3}_v$, which ensures the soundness of treating labelled tensor as an AC symbol.

@@

%Define $T_{s,s'}$ as $\SType$ if $s=s'=\emptyset$, $\KType(\sigma_s)$ if $s'=\emptyset$, $\BType(\sigma_{s'})$ if $s=\emptyset$ and $\OType(\sigma_s,\sigma_{s'})$ otherwise (see Sec. \ref{context: sigma s}). We claim the soundness of our normalization as:
% Then if the original equality is $\Gamma\vdash D_1 = D_2 : \cD(s,s')$, what remains to be check $\Gamma\vdash e_1 = e_2 : T_{s,s'}$.

By routinely showing that Lemma \ref{lem: axiom sound} holds for the extra rules mentioned above and in Appendix, we further claim:
\begin{theorem}[Soundness and Completeness of Normalization]
  Assuming $\Gamma\vdash D_1 : \cD(s,s')$, $\Gamma\vdash D_2 : \cD(s,s')$, 
  and no variables of $\DType(\cdot,\cdot)$ in $D_1,D_2$.
  Let $e_1 = e_2$ be obtained by above normalization procedure on $D_1 = D_2$. Then $\Gamma\vDash D_1 = D_2$ if and only if $\Gamma\vDash e_1 = e_2$.
  %All rewriting rules for labelled Dirac notations listed here and in Appendix are sound, i.e., for any interpretation $\sem{\cdot}$, $D\reduce D'$ implies $\sem{D} = \sem{D'}$.
\end{theorem}
% \begin{proof}
%   Notice that all rewriting rules and ordering perserve the types, so the final step eliminating labels is both sound and complete. 
%   It is routinely to check that each rewriting rules and ordering step are sound based on the semantics. This directly leads to the soundness. For completeness, we need further check if the normalization procedure always provide $e_1 = e_2$, i.e., it never gets stuck and terminates. Step 1 is trivial. For step 2, it can be further checked that 1. there is no circular rules and thus terminates, and 2. if it is not in form (\ref{eqn:Labelled normal}), there must exists one of the rules can be further applied. For step 3, ordering is always applicable. This completes the proof.
% \end{proof}
